Davos: a Python package ``smuggler'' for constructing lightweight reproducible notebooks

### 2023-09-30 paper/main.tex
big update to impact section

Like virtual environments, containers, and virtual machines, the
Davos package (when used in conjunction with Jupyter
notebooks) provides a light\-weight mechanism for sharing code and
ensuring reproducibility across users and computing environments
(Fig.~\ref{fig:code-sharing}). Further, Davos enables users
to fully specify (and install, as needed) any project dependencies
within the same notebook. This provides a system whereby executable
code (along with text and media) \textit{and} code for setting up and
configuring the project dependencies, may be combined within a single
notebook file.

Although existing notebooks \textit{can} incorporate system calls that install
project requirements, handling project requirements in the general case is
non-trivial (e.g., see Fig.~\ref{fig:flow-chart}). Further, Davos
incorporates its own virtual environment system that isolates
notebook-installed packages from the runtime environment
(Sec.~\ref{subsec:projects}). In many setups this feature can eliminate the
need to set up a separate virtual environment or container (e.g., in
conjunction with a \texttt{requirements.txt}, \texttt{project.toml}, or
\texttt{environment.yml} file specifying the project's dependencies).

We designed Davos for use in research applications. For
example, in many settings, Davos may be used as a drop-in
replacement for more-difficult-to-set-up virtual environments,
containers, and/or virtual machines. For researchers, this lowers
barriers to sharing code. By eliminating most of the setup costs of
reconstructing the original researchers' computing environment,
Davos also lowers barriers to entry for members of the
scientific community and the public who seek to run shared code.

>>>>

We designed Davos for use in research settings, where code for numerous different tasks---from processing data, to running statistical analyses, to generating figures and tables for publication---is frequently shared among collaborators while working on a project, and eventually with the broader scientific community and general public upon a its completion.
In these contexts, ensuring that shared code yields consistent, reproducible outputs across users and over time is crucially important, yet the tools available to researchers for doing so can be complex to set up and challenging to properly use.
This has the dual effect of discouraging scientists from sharing their code in a reproducible way (or at all), and making it significantly harder (or impossible) for others to successfully reproduce their results, adopt or extend their methods, and contribute to or build upon their work.
Ultimately, the need to install and master additional tools in order to share and run reproducible code can impede progress both on the individual level and on the broader scientific level.

While the Davos package by no means offers a universal solution to this problem, it can, in many cases, provide an effective yet more accessible alternative to existing systems for sharing reproducible Python code, such as virtual environments, containers, and virtual machines (see Fig.~\ref{fig:code-sharing}).
For researchers, this can lower barriers to collaborating with peers and contributing to publicly available open science resources.
And by eliminating most of the setup costs of reconstructing the original researchers' computing environment, Davos also lowers barriers to entry for members of the scientific community and the public who seek to run shared code.

Among common systems for sharing reproducible Python code (see Fig.~\ref{fig:code-sharing}), Davos is most comparable to virtual environments as it similarly provides a lightweight mechanism for specifying and sharing a complete set of packages (and specific package versions) required by a particular project, and installing those packages into an isolated directory on the user's file system.
However, when used in conjunction with Jupyter notebooks, Davos offers a number of advantages over standard virtual environments that make it both easier to use and more effective at ensuring reproducibility of shared code.

First, \texttt{smuggle} statements and onion comments enable researchers to specify their notebooks' dependencies directly within the code that requires them (see Secs.~\ref{subsec:smuggle} and  \ref{subsec:onion}).
This eliminates the need to either manually create and maintain separate configuration files or use an additional tool to generate them, and to then distribute these files alongside their primary code base for users to download.

Second, Davos automatically checks for, installs, isolates, and imports any required packages at runtime.
This allows users to download and immediately run a Davos-enhanced notebook without any prerequisite setup (beyond that needed to run a standard Jupyter notebook).
By contrast, before running a notebook whose dependencies are managed via a virtual environment, the user would first need to run a series of shell commands to manually create a new environment, populate it with packages from the researcher's configuration file (which the user must also have obtained), and then either use the \texttt{ipykernel} package to register the environment as a Jupyter kernel, or activate and deactivate the environment before and after (respectively) each time the notebook was launched.
Beyond reducing this initial setup cost, Davos's runtime-based approach to dependency management affords a second important benefit.
While creating and configuring a virtual environment ensures that a specific set of packages is initially installed, it does not guarantee that they will \textit{remain} installed after that point.
For example, a researcher who creates a virtual environment in which to run a set of data analyses---or a different user who later recreates that environment to reproduce them---might at some point install an additional package into the environment after its initial setup (e.g., to implement a new analysis idea).
Depending on the requirements of this new package, this could cause one or more initially installed packages to be upgraded or downgraded to a different version.
If the individual does not happen to notice this change when it occurs, differences between those packages' expected and installed versions may introduce bugs into previously written code or subtly alter its output when it is next run.
Because \texttt{smuggle} statements and onion comments are evaluated every time a Davos-enhanced notebook is run, they function to ensure that the notebook's dependencies are always satisfied and that any such inadvertent changes would be automatically caught and corrected.

Third, running a shared notebook that uses Davos to manage its dependencies often requires less (but never more) space on the user's system than running an identical notebook inside a virtual environment.
While typical virtual environment directories will contain \textit{all} requirements listed in their configuration files, Davos's isolated project directories (by default; see Sec.~\ref{subsec:projects}) contain only those not already available in the user's existing Python environment.
This is another feature made possible by Davos's runtime-based dependency management scheme.
Davos can safely draw from packages in from the user's main environment to satisfy a notebook's dependencies, when possible, because if those packages were to be updated or removed at any point such that they no longer met the notebook's requirements, appropriate versions could (and would) be installed into the project directory the next time the notebook was executed.

Finally, neither virtual environments nor Davos-enhanced notebooks can bundle specific versions of Python with which to run shared code.
However, Davos's \texttt{require\_\-python()} function provides a simple mechanism for indicating a required or constrained Python version and alerting the user at runtime (by raising an exception) if their Python version is incompatible.
In terms of Davos's ability to ensure that shared code can be executed reproducibly by other users, this falls short of the capabilities of more complex tools that can provide complete copies of Python (see Fig.~\ref{fig:code-sharing}) and should be preferred over Davos in cases where this is necessary.
However, the functionality it provides over what is possible with a standard virtual environment is removing the \textit{expectation} that running a particular notebook will reproduce an expected outcome in situations where this is either impossible or not guaranteed.
With this understanding, a user may choose to install a compatible Python version through some other means or elect to still run the code, but will not be surprised by a potential failure to execute successfully or output an expected result.
